---
title: "What is Data Science?"
output:
  html_document: 
    toc: true
    toc_float: true
---

We have, by now, established some fundamental tools for doing data science. It's important to revisit our definition, and especially our discussion of connotation, before moving forward.

```{r, echo = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

library(tidyverse)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```

## Overview {.tabset .tabset-pills}

### Learning Objectives

Pull together skills learned through this point to produce analytical summaries and reports.

### Slide Deck

<div class="vid_container">
  <iframe 
    src="https://speakerdeck.com/player/69791e0a56b14efd9f741fc10cea30f9" 
    allowfullscreen 
    frameborder="0"
    class="video">
  </iframe>
</div>

<div style="margin-bottom:5px"> <strong> <a href="https://speakerdeck.com/jeffgoldsmith/p8105-data-science-revisited" title="Writing with data" target="_blank">What is Data Science? Revisited</a> </strong> from <strong><a href="https://speakerdeck.com/jeffgoldsmith" target="_blank">Jeff Goldsmith</a></strong>. </div><br>

***


## Example

This example is based entirely on live-coding and uses the [NYC Airbnb data](dataset_airbnb.html). The data can be imported using the [p8105.datasets](https://github.com/P8105/p8105.datasets) package:

```{r}
library(p8105.datasets)

data(nyc_airbnb)
```

As always, I'll do today's coding in a R Markdown file, sitting in an GitHub Repo / R Project.

### Understanding variables

First, let's take a few minutes to understand the dataset and the variables it contains.

```{r}
# View(nyc_airbnb)
str(nyc_airbnb)

nyc_airbnb %>%
  count(room_type)

nyc_airbnb %>%
  count(neighbourhood_group)
```

### Brainstorming questions

A major element of data science is to ask questions, and this dataset provides some rich opportunities. For example, we might ask:

* Does rating vary by neighborhood, room type, or both?
* How is price related to other variables?
* Where are rentals located?

We'll take a few minutes as a class to brainstorm some additional questions, and then try to answer some of them.

```{r, echo = FALSE, eval = FALSE}
ggplot(nyc_airbnb, aes(x = longitude, y = latitude)) +
  geom_point() + 
  coord_map()
  coord_equal(ratio = 1)
```

```{r, echo = FALSE, eval = FALSE}
library(leaflet)
library(p8105.datasets)

data("nyc_airbnb")

nyc_airbnb = 
  nyc_airbnb %>% 
  mutate(stars = review_scores_location / 2) %>% 
  rename(boro = neighbourhood_group)

pal <- colorNumeric(
  palette = "viridis",
  domain = nyc_airbnb$stars)

nyc_airbnb %>% 
  filter(boro %in% c("Manhattan", "Brooklyn", "Queens")) %>% 
  na.omit(stars) %>% 
  sample_n(5000) %>% 
  mutate(click_label = str_c("<b>$", price, "</b><br>", stars, " stars<br>", number_of_reviews, " reviews")) %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~lat, ~long, radius = .1, color = ~pal(stars), popup = ~click_label)
```

## Other materials

Some additional links on `leaflet`:

* A quick introduction is [here](https://teachdatascience.com/leaflet/)
* A more thorough overview is [here](https://rstudio.github.io/leaflet/)

Some other reading on thinking like a data scientist:

* If you didn't listen before, now's the time for Chris Volinsky's ["How Industry Views Data Science Education in Statistics Departments"](https://youtu.be/JLs01Z5baSU?t=1063)
* Jeff Leek's [problem forward](https://simplystatistics.org/2013/05/29/what-statistics-should-do-about-big-data-problem-forward-not-solution-backward/) blog post
* This (somewhat long) [post](https://simplystatistics.org/2019/01/09/how-data-scientists-think-a-mini-case-study/) has great insights into a data science mindset 
* Although we're playing with public (scraped) data, you might be interested in [How R Helps Airbnb Make the Most of Its Data](https://peerj.com/preprints/3182/)

The code that I produced working examples in lecture is [here](https://github.com/P8105/case_study).