---
title: "NYC Airbnb"
author: "Jeff Goldsmith"
date: "10/9/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
```

## Data import

I'll load the tidyverse and the dataset. I'm also doing a bit of editing and renaming to make my life easier. 

```{r}

```

## Understanding variables

Next I'll do a bit of inspection to make sure I understand the data structure.

```{r}

```

## Brainstorming questions

### Does rating vary by neighborhood, room type, or both?


### Do hosts with more listings have higher or lower ratings?


### Where are rentals located?


### Other questions

There are several questions we didn't get to:

* How is price related to other variables?
* How are descriptions related to number of reviews?
* Which rentals make the most money?
* What's different about low-availability rooms?

Of these, the question dealing with descriptions is one we're not in a position to handle yet -- that involves parsing character strings, which we'll get to shortly. The others we could at least explore using the tools we have at our disposal now.


## Discussion

Need a discussion.